Does personal relevance attenuate the effect of government branding? Evidence from a survey experiment on air-pollution policy

### 2019-07-19 paper3_relevance and symbols_revision_01.Rmd
Add files via upload

---
title: "Does personal relevance attenuate the effect of government branding? Evidence from a survey experiment on air-pollution policy"
author: '**Saar Alon-Barkat**'
date: " "
output:
  html_document:
    theme: flatly
    toc: yes
    toc_float:
      collapsed: no
      smooth_scroll: yes
      toc_depth: 3
  pdf_document:
    toc: yes
  word_document: default
link-citations: yes
bibliography: phd_paper_2.bib
csl: https://raw.githubusercontent.com/citation-style-language/styles/master/apa.csl
urlcolor: blue
---

@@

Public Administration scholarship has recently shown a growing interest in public sector branding, and its potency to influence citizens' perceptions and direct their behaviour [@eshuis_2012; @marland_2016; @marland_2017; @zavattaro_website_2018; @sataen_2015]. Behavioural studies established that symbolic, brand elements[^footnote_branding] entangled in government communications (e.g. brand names, logos, colors and figures) can have an emotional effect on citizens, which leads them to view public organizations and their policies and services more favourably [@marvel_2015a; @karens_2016; @alonbarkat_2017; @teodoro_2018]. This growing scholarly interest corresponds with the overall increase in the use of branding practices by public-sector organizations over the past decade, in part due to the expension of social media and ditigal communication platforms [@eshuis_2012; @marland_2017; @mickoleit_2014; @mergel_2013].

@@

The ability of government organizations to affect citizens via symbolic communications can be perceived favorably, insofar as it enables organizations to engage more effectively with citizens and overcome their unjustified negative prior beliefs and biases against the public sector [@marvel_2015a; @marvel_2015; @hvidman_2016; @hvidman_2019]. Accordingly, it can mitigate citizens' distrust in government organizations and underevaluation of their performance, which represent a serious concern.

Nonetheless, the persuasion of citizens through symbols is also highly problematic. Normatively, we expect citizens in a democratic society to form their opinions about their government based on their critical thinking, rather then on uncontious emotive responses, produced by government agencies themselves. A yet more serious concern regards the potency of symbolic elements in government communications to compensate for organizations' poor performance, poorly planned policy plans and logically unpersuasive explanations, as indicated by previous studies [@alonbarkat_2017; @alonbarkat_2018]. They can yield a misalignment between citizens' positive perceptions and the "real world", which would undermine the democratic accountability and responsiveness of public organizations. Moreover, government organizations (or politicians controlling them) could strategically exploit such symbolic communication to manipulate public opinion and mitigate justified public criticism of agencies' failures.

The abovementioned concerns of the negative implications of branding are the main motivation of this study. These concerns necessitate a better understanding of the boundaries of governments' persuasion (and possibly manipulation) of citizens through symbolic communication. Understanding the limitations of symbolic persuasion would enable to asses the scope of the risk that these practices pose for democracy.

@@

# The persuasive effect of symbols and its moderation by the perceived personal relevance of the communication.

Symbolic elements, that evoke particular associations and emotions, are included in almost every government communication with citizens, and most most prominently in government advertising and marketing [@eshuis_2012]. These symbols can affect citizens' attitudes unconsciously through a psychological mechanism of evaluative conditioning -- the positive associations towards the symbolic element is transferred to the public organization, following repeated appearance of the symbol in the organization's public communications [@alonbarkat_2017; @dehouwer_2012]. This mechanism is considerred a primitive form of persuasion which requires very little cognitive effort, as opposed to persuasion resulted by thoughtful consideration of substantive arguments (in the administrative context, for instance - evaluating organizations and policies based on performance information). The question under which circumbstances are citizens' more/less susceptible to persuasion via branding, falls within a more general theoretical question of what determines the ways in which individuals process communication messeges.

According to the Elaboration Likelihood Model (ELM), developed in the late 1970s by social psychology scholars Richard Petty and John Cacioppo [@petty_1986; @petty_2011], people may use different processing mechanisms in response to a communication, and accordingly persuasion can occure in different ways.  
depending on the amount of thinking that they dedicate to it. When people have high motivation and ability to elaborate on a message, they are likely to focus on the arguments in the communication that are central to the merits of the issue, and scrutinize them ("central route"). Conversely, when they are unmotivated or unable to elaborate, they may rely on available heuristic cues that require relatively little amount of cognitive effort ("peripheral route"). These two routes for persuasion are regarded as two ends of an "elaboration continuum". People may also be located between these two ends of the continuum, and then a mixture of these processes will determine their persuasion.

Viewed against the above theoretical framework, symbolic elements can be regarded as peripheral cues, which can affect attitudes by evoking emotions and associations attributed to government organizations and their policies (peripheral route). Accordingly, the more citizens are motivated to think about the content of the communication, and the more capable they are of doing so, the more they are likely to form their attitudes based on elaborate processing about the substantive content, and the less susceptible they are to symbolic elements in the communication.

This literature has pointed to a variety of factors that shape people's motivation and ability to think about a message, and accordingly affect their inclination to be persuaded either by the arguments or by peripheral cues. Among others, scholars have theorized that the more people perceive an issue as personally relevant, the more motivated they are to invest in thinking about it, since the consequences of being incorrect are greater. This postulation has been empirically tested, and largely confirmed, by numerous ELM based experimental studies in social psychology and marketing [@petty_1979; @petty_1981; @petty_1983; @johnson_1989; @petty_1990; @burnkrant_1989]. These studies have demonstrated that increasing the perceived personal relevance of a communication attenuates the effect of peripheral cues. For intance, petty et al. 1983 
Later reviews of ELM literature considered the factor of perceived personal relevance as the most important determinant of motivation to process a message [@petty_2002; @petty_2011].[^footnote_elm_factors] ## add quote ##

@@

Connecting the above public administration literature on citizens' responses to government public communications with the social psychology ELM literature, I theorize that:

*H<sub>1</sub> - Symbolic elements in communications are less (more) likely to increase citizens' trust in government organizations and policies, the more (less) citizens perceive the policy issue in the communication as having personal relevance for themselves.*

@@

# The moderating effect of perceived personal relevance on argument quality.

@@

*H<sub>2</sub> - The logical quality of the communication is more (less) likely to affect citizens' trust in government organizations and policies, the more (less) citizens perceive the policy issue in the communication as having personal relevance for themselves.*

>>>>

---
title: "Does personal relevance attenuate the effect of government branding? Evidence from a survey experiment on air-pollution policy"
author: '**Saar Alon-Barkat**'
date: " "
always_allow_html: yes
output:
  html_document:
    theme: flatly
    toc: yes
    toc_float:
      collapsed: no
      smooth_scroll: yes
      toc_depth: 3
  pdf_document:
    toc: yes
  word_document: default
link-citations: yes
bibliography: phd_paper_2.bib
csl: https://raw.githubusercontent.com/citation-style-language/styles/master/apa.csl
urlcolor: blue
---

@@

Public Administration scholarship has recently shown a growing interest in public sector branding, and its potency to influence citizens' perceptions and direct their behaviour [@eshuis_2012; @marland_2016; @marland_2017; @zavattaro_website_2018; @sataen_2015]. Behavioural studies established that symbolic, brand elements[^footnote_branding] entangled in government communications (e.g. brand names, logos, colors and figures) can have an emotional effect on citizens, which leads them to view public organizations and their policies and services more favourably [@marvel_2015a; @karens_2016; @alonbarkat_2017; @teodoro_2018]. This growing scholarly interest corresponds with the overall increase in the use of branding practices by public-sector organizations over the past decade, in part due to the expension of ditigal communication platforms and the use of social media [@eshuis_2012; @marland_2017; @mickoleit_2014; @mergel_2013].

@@

The ability of government organizations to affect citizens via symbolic communications can be perceived favorably, insofar as it enables organizations to engage more effectively with citizens and overcome their unjustified negative prior beliefs and biases against the public sector [@marvel_2015a; @marvel_2015; @hvidman_2016; @hvidman_2019]. Accordingly, it can mitigate citizens' distrust in government organizations and underevaluation of their performance.

Nonetheless, the persuasion of citizens through symbols is also highly problematic. Normatively, we expect citizens in a democratic society to form their opinions about their government based on their critical thinking, rather then on uncontious emotive responses, produced by government agencies themselves. A yet more serious concern regards the potency of symbolic elements in government communications to compensate for organizations' poor performance, poorly planned policy plans and logically unpersuasive explanations, as indicated by previous studies [@alonbarkat_2017; @alonbarkat_2018]. Congruently, government branding can yield a misalignment between citizens' positive perceptions and the "real world", which would undermine the democratic accountability and responsiveness of public organizations. Moreover, government agencies (or politicians controlling them) could strategically use such symbolic communication as a propaganda, seeking to manipulate public opinion and mitigate justified public criticism of agencies' failures.

The abovementioned concerns of the negative implications of branding are the main motivation of this study. These concerns necessitate a better understanding of the boundaries of governments' persuasion (and possibly manipulation) of citizens through symbolic communication. Understanding the limitations of symbolic persuasion would enable to asses more effectively the scope of the risk that these practices pose for democracy.

@@

# The effect of government branding on citizens' attitudes and its moderation by the perceived personal relevance of the communication.

Symbolic elements that are strategically selected and designed to evoke positive emotions and associations, are included in almost every government communication with citizens, and most most prominently in government advertising and marketing [@eshuis_2012]. These symbols can affect citizens' attitudes unconsciously through a psychological mechanism of evaluative conditioning, whereby the positive associations triggerred by the symbolic element are transferred to the public organization [@alonbarkat_2017; @dehouwer_2012]. So for instance, when a public organization uses a popular celebrity endorser in a campaign, the positive connotations for the celebrity (e.g. physically attractive, funny, talented) can be extended to the organization, which would lead citizens to also view the organization more favorably. This mechanism is considerred a primitive form of persuasion which requires very little cognitive effort, as opposed to persuasion resulted by thoughtful consideration of substantive arguments. Thus, the question of under which circumbstances are citizens' more/less susceptible to persuasion via branding, is linked to a more theoretical discussion in social psychology regarding the relations between different psychological processes of persuasion or modification of people's attitudes or judgements in response to communications or messages, and the determinants of these different processes.

According to the Elaboration Likelihood Model (ELM) [@petty_1986; @petty_2011], persuasion can occur via different process that can be placed along an "elaboration continuum". At the high end of that continuum, also called "central route", persuasion occurs through processes that require high degrees of thought and cognitive effort. This type of processing involves thoughtful scrutiny of arguments in the communication that are central to the merits of the issue. People located at the high end are more likely to form their judgements on the matter based on the logical quality of the arguments, and are more likely to be persuaded by strong, logically persuasive arguments. As for those who are located at the low end of that continuum, also called "peripheral route", persuasion may occur mainly through processes that require little ammount of thinking, including evaluative conditioning, and the reliance on other heuristic cues like the message source [@james_2017a]. When people are located between these two ends of the continuum, a mixture of these processes will operate and determine their persuasion, depending on their relative position.

In any given situation, the position of people along that continuum would be determined, according to ELM, by their motivation and ability to scrutinize the message. The more motivated people are to elaborate on the message and the more capable they are of doing so, the more likely they are to focus on the substative arguments, and likewise the less susceptible they are to persuasion via peripheral route processes.

ELM literature has pointed to a variety of factors that shape people's motivation and ability to think about a message, and accordingly affect their inclination to be persuaded either by the arguments or by peripheral cues. Perhaps the most studied factor in this regard is the perceived personal relevance of the communication (or "issue involvement"), namely the extent to which they perceive the issue at hand as having significant consequences for their own lives. That variable has been considered the most important determinant of motivation to scrutinize the message [@petty_2002; @petty_2011].[^footnote_elm_factors] It has been theorized that the more people perceive an issue as personally relevant, the more motivated they are to invest in thinking about it, since the consequences of being incorrect are greater [@petty_1979; @petty_1981; @petty_1983; @johnson_1989; @petty_1990; @burnkrant_1989]. Studies have established that individuals are more affected by peripheral cues and less affected by argument quality, the less they perceive the issue as personally relevant. For instance, in an experimental marketing study by @petty_1983, participants were asked to express their attitudes about a new product (disposable razor), after being exposed to a magazine ad, which contained either strong or weak arguments about
the product, and involved either a non-famous endorser or a celebrity endorser (a symbolic peripheral cue). To manipulate the personal relevance of the product, subjects were either told that the product would soon be available in their local area (high relevance), or in other areas (low relevance). The results showed the effect of the celebrity endorser was higher among the low personal relevance group, whereas the effect of argument quality was significantly weaker for that group.

@@

Viewed against the above theoretical framework, symbolic elements can be regarded as peripheral cues, which can affect attitudes by a peripheral route process of evaluative conditioning.[##add footnote, in rare occations, symbolic elements could also serve as arguments, e.g. a slogan that both has a symbolic meaning and also contains an argument##]. Hence, ELM postulates that citizens' are more likely to be affected by symbols the less motivated and capable they are to critically think about the message, which likely to be determined by the perceived personal relevnace of the policy issue.

*H<sub>1</sub> - Symbolic elements in communications are likely to have a greater positive effect on citizens' attitudes about government organizations and policies, the more citizens perceive the policy issue in the communication as having personal relevance for themselves.*

*H<sub>2</sub> - The logical quality of the substantive arguments in the communication is more likely to affect citizens' attitudes about government organizations and policies, the more citizens perceive the policy issue in the communication as having personal relevance for themselves.*

@@

# The effect of substantive information in moderating effect of perceived personal relevance on argument quality.

@@
